---
title: "Lab 02: Matrix Decompositions"
author: 
- "Donggyun Kim"
- "27008257"
date: "1/31/2018"
output: pdf_document
---

# 1) Singular Value Decomposition
\[
\mathbf{M} = \mathbf{UDV^\top}
\]

- $\mathbf{M}$: $n \times p$ matrix of full column-rank $p$, $(n > p)$  
- $\mathbf{U}$: $n \times p$ matrix of left singular vectors  
- $\mathbf{D}$: $p \times p$ diagonal matrix of singular values  
- $\mathbf{V}$: $p \times p$ matrix of right sungular vectors  
  
  
1) Use `svd()` to compute SVD of state.x77
```{r}
SVD <- svd(state.x77)
SVD
```
  
  
2) Create the matrices, $\mathbf{U}$, $\mathbf{D}$, and $\mathbf{V}$
```{r}
U <- SVD$u
U

D <- diag(SVD$d)
D

V <- SVD$v
V
```
  
  
3) Confirm that the data state.x77 can be obtained as the product of: $\mathbf{UDV^\top}$
```{r}
UDtV <- U %*% D %*% t(V)

head(UDtV, 10)

head(state.x77, 10)
```

**SVD and best Rank-one Approximations**
\[
\mathbf{X_r} = \displaystyle \sum_{k = 1}^r l_k\mathbf{u}_k{\mathbf{v}_k}^t
\]
1) Create an object state2 by selecting the first five columns of state.x77
```{r}
state2 <- state.x77[, 1:5]
head(state2, 10)
```
  
  
2) Perform the SVD decomposition of state2
```{r}
SVD_2 <- svd(state2)
SVD_2
```
  
  
3) Confirm that the sum: $\displaystyle\sum_{k=1}^5l_k\mathbf{u}_k\mathbf{v}_k^t$ equals state2
```{r}
l <- SVD_2$d
U <- SVD_2$u
V <- SVD_2$v

X <- matrix(0, nrow = nrow(state2), ncol = ncol(state2))
for (i in 1:length(l)) {
  X <- X + l[i] * U[, i] %*% t(V[, i])
}

head(X, 10)

head(state2, 10)
```


**Using SVD output to visualize data**
```{r}
library(ggplot2)

U <- svd(state.x77)$u
U1 <- U[, 1]
U2 <- U[, 2]

plot(U1, U2, main = "Plot of States (first 2 left singular vectors")
text(U1, U2, rownames(state.x77))
```

```{r}
V <- svd(state.x77)$v

V1 <- V[, 1]
V2 <- V[, 2]

plot(V1, V2, main = "Plot of Variables (first 2 right singular vectors")
text(V1, V2, colnames(state.x77))
```

# 2) Eigenvalue Decomposition
1) Compute a matrix X as the mean-centerd data of state.x77
```{r}
X <- sweep(state.x77, 2, colMeans(state.x77), "-")
head(X, 10)
```

2) Calculate the sum-of-sqares and cross-products matrix $\mathbf{S} = \mathbf{X}^\top\mathbf{X}$
```{r}
S <- crossprod(as.matrix(X))
S
```

3) Confirm that cov($\mathbf{X}$) is equal to $\frac{1}{n - 1}\mathbf{X}^\top\mathbf{X}$
```{r}
n <- nrow(state.x77)
S / (n - 1)

cov(X)
```

4) Use `solve()` to compute the inverse $\mathbf{S}^{-1}$
```{r}
solve(S)
```

5) USE `eigent()` to compute the EVD of $\mathbf{S} = \mathbf{V \Lambda V}^\top$
```{r}
EVD <- eigen(S)
EVD
```

6) Confirm that $\mathbf{S}^{-1}$ can also be obtained as: $\mathbf{V}\mathbf{\Lambda}^{-1}\mathbf{V}^\top$
```{r}
V <- EVD$vectors
L <- diag(EVD$values)

V %*% solve(L) %*% t(V)

solve(S)
```

**Power Method**
```{r}
A <- matrix(c(5, -14, 11, -4, 4, -4, 3, 6, -3), nrow = 3, byrow = 1)
w0 <- c(1, 1, 1)

eig <- function(X, v, k) {
  for (i in 1:k) {
    v_p <- X %*% v
    v_c <- v_p / max(abs(v_p))
    v <- v_c
  }
  val <- as.numeric(crossprod(v, X) %*% v / crossprod(v, v))
  eig <- list(val, v)
  names(eig) <- c("value", "vector")
  eig
}
  
eig(A, w0, 10)

eigen(A)
```

**$l_p$ norm for Power Method**
```{r}
eig <- function(X, v, k, norm = Inf) {
  for (i in 1:k) {
    v_p <- X %*% v
    if (norm == Inf) {
      nm <- max(abs(v_p))
    } else {
      nm <- (sum(abs(v_p)^norm))^(1/norm)
    }
    v_c <- v_p / nm
    v <- v_c
  }
  val <- as.numeric(crossprod(v, X) %*% v / crossprod(v, v))
  eig <- list(val, v)
  names(eig) <- c("value", "vector")
  eig
}

eig(A, w0, 20)  # l-infinity norm

eig(A, w0, 20, 2)  # l-2 norm

eigen(A)
```

**Deflation and more eigenvectors**
```{r}
deflation <- function(X, v, k, norm = Inf) {
  
  eig <- eig(X, v, k, norm)
  def <- list(eig$value, eig$vector)
  names(def) <- c("values", "vectors")
  
  for(i in 2:nrow(X)) {
  X <- X - eig$value * tcrossprod(eig$vector)
  eig <- eig(X, v, k, norm)
  def$values <- c(def$values, eig$value)
  def$vectors <- cbind(def$vectors, eig$vector)
  }
  def
}

deflation(A, w0, 20, 2)

eigen(A)
```