---
title: "Lab 01: Basics of Matrix Algebra in R"
author: "Donggyun Kim"
date: "1/24/2018"
output: html_document
---

# 1) Basic Vector and Matrix manipulation in R

A vector $\mathbf{x}$
```{r}
x <- 1:9
```

Using $\mathbf{x}$, create following matrices
```{r}
matrix(x, nrow = 3, ncol = 3)
matrix(x, nrow = 3, ncol = 3, byrow = 1)
```

Create identity matrix
```{r}
diag(5)
```

Three vectors, $\mathbf{a_1}, \mathbf{a_2}, \mathbf{a_3}$
```{r}
a1 <- c(2, 3, 6, 7, 10)
a2 <- c(1.88, 2.05, 1.70, 1.60, 1.78)
a3 <- c(80, 90, 70, 50, 75)
```

Use `cbind()` to create a matrix, $A_{5 \times 3}$
```{r}
A <- cbind(a1, a2, a3)
rownames(A) <- 1:5
A
```

Three vectors, $\mathbf{b_1}, \mathbf{b_2}, \mathbf{b_3}$
```{r}
b1 <- c(1, 4, 5, 8, 9)
b2 <- c(1.22, 1.05, 3.60, 0.40, 2.54)
b3 <- c(20, 40, 30, 80, 100)
```

Use `rbind()` to create a matrix, $B_{3 \times 5}$
```{r}
B <- rbind(b1, b2, b3)
colnames(B) <- 1:5
B
```

Compute matrix products using `%*%` operator  
$\mathbf{AB}$
```{r}
A %*% B
```
$\mathbf{BA}$
```{r}
B %*% A
```
$\mathbf{A^\top B^\top}$
```{r}
t(A) %*% t(B)
```
$\mathbf{B^\top A^\top}$
```{r}
t(B) %*% t(A)
```

Obtain a linear combination via a matrix multiplication  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; 1 $\times$ Sepal.Length $+$ 2 $\times$ Sepal.Width $+$ 3 $\times$ Petal.Length $+$ 4 $\times$ Petal.Width  
```{r}
M_iris <- as.matrix(iris[1:4])
N <- matrix(rep(1:4, nrow(iris)), nrow = 4)
M <- M_iris %*% 1:4
rownames(M) <- 1:150
colnames(M) <- paste(1:4, colnames(M_iris), collapse = " + ")
head(M, 10)
```

Write a function `vnorm()` that computes the length of a vector:  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $||\mathbf{v}|| = \sqrt{\mathbf{v^\top}\mathbf{v}}$
```{r}
vnorm <- function(v) {
  as.numeric(sqrt(t(v) %*% v))
}
```

Given a vector, $\mathbf{v}$
```{r}
v <- 1:5
```

Find a unit vector of $\mathbf{v}$
```{r}
u <- v / vnorm(v)
u
```

The length of $\mathbf{u}$ is
```{r}
vnorm(u)
```

Write a function `is_square()` to check whether the provided matrix is a square matrix
```{r}
is_square <- function(X) {
  if (!is.matrix(X)) {
    stop("The input must be a matrix")
  }
  
  if (nrow(X) == ncol(X)) {
    TRUE
  } else {
    FALSE
  }
}
```

Test if the function works
```{r}
D <- diag(4)
D
is_square(D)

A
is_square(A)
```

Write a function `mtrace()` to compute the trace of a square matrix
```{r}
mtrace <- function(X) {
  if (!is.matrix(X) | !is_square(X)) {
    stop("The input must be a suare matrix")
  }
  
  sum(diag(X))
}
```

Given two square matrices $\mathbf{A}$ and $\mathbf{B}$, verify that `mtrace()` is linear mapping:
```{r}
A <- matrix(1:9, ncol = 3, nrow = 3)
A

B <- diag(3)
B
```

- $\operatorname{tr}(A + B) = \operatorname{tr}(A) + \operatorname{tr}(B)$
```{r}
mtrace(A + B)
mtrace(A) + mtrace(B)
```

- $\operatorname{tr}(cA) = c \times \operatorname{tr}(A)$
```{r}
mtrace(3 * A)
3 * mtrace(A)
```

**Trace of products**  
Given two matrices $\mathbf{X}$ and $\mathbf{Y}$, verify that
$$
\operatorname{tr}(\mathbf{X^\top Y}) = \operatorname{tr}(\mathbf{XY^\top}) = \operatorname{tr}(\mathbf{Y^\top X}) = \operatorname{tr}(\mathbf{YX^\top})
$$

- Prove that trace function is commutative  
$$
\operatorname{tr}(\mathbf{AB}) = \operatorname{tr}(\mathbf{BA})
$$
Because $\mathbf{AB}_{ij} = \displaystyle \sum_{k = 1}^{n}a_{ik}b_{kj}$ for matrices $\mathbf{A_{p \times n}}$ and $\mathbf{B_{n \times p}}$,   
$\operatorname{tr}(\mathbf{AB}) = \displaystyle \sum_{i = 1}^p\sum_{k = 1}^na_{ik}b_{ki}$ and $\operatorname{tr}(\mathbf{BA}) = \displaystyle \sum_{k = 1}^n\sum_{i = 1}^pb_{ki}a_{ik}$.  
$$
\operatorname{tr}(\mathbf{AB}) = \displaystyle \sum_{i = 1}^p\sum_{k = 1}^na_{ik}b_{ki} = \displaystyle \sum_{i = 1}^p\sum_{k = 1}^nb_{ki}a_{ik} = \displaystyle \sum_{k = 1}^n\sum_{i = 1}^pb_{ki}a_{ik} = \operatorname{tr}(\mathbf{BA})
$$

- Prove that the trace of a square mtrix, $\mathbf{A}_{n \times n}$ equals the trace of transpose of the square matrix
$$
\operatorname{tr}(\mathbf{A}) = \operatorname{tr(\mathbf{A^\top)}}
$$
Because $\mathbf{A}_{ij} = a_{ij}$ and $\mathbf{A^\top}_{ij} = a_{ji}$,  
$\operatorname{tr}(\mathbf{A}) = \displaystyle \sum_{i = 1}^na_{ii}$ and $\operatorname{tr}(\mathbf{A^\top}) = \displaystyle \sum_{j = 1}^na_{jj}$  
$$
\operatorname{tr}(\mathbf{A}) = \displaystyle \sum_{i = 1}^na_{ii} = \displaystyle \sum_{j = 1}^na_{jj} = \operatorname{tr}(\mathbf{A^\top}) 
$$

It is proved that $\operatorname{tr}(\mathbf{XY^\top}) = \operatorname{tr}(\mathbf{Y^\top X)}$


Let $\mathbf{P = X^\top Y}$. This implies $\mathbf{P^\top} = (\mathbf{X^\top Y})^\top = \mathbf{Y^\top X}$   
Also, $\operatorname{tr}(\mathbf{P}) = \operatorname{tr}(\mathbf{X^\top Y}) = \operatorname{tr}(\mathbf{Y^\top X}) = \operatorname{tr}(\mathbf{P^\top})$  
  
  
Using two properties of trace function above, it is verified that 
$$
\operatorname{tr}(\mathbf{X^\top Y}) = \operatorname{tr}(\mathbf{XY^\top}) = \operatorname{tr}(\mathbf{Y^\top X}) = \operatorname{tr}(\mathbf{YX^\top})
$$

---

# 2) Transformation and Scaling Operations

Create a Matrix $\mathbf{M}$
```{r}
M <- mtcars[c("mpg", "disp", "hp", "drat", "wt")]
head(M, n = 10)
```

Compute the vector containing the means of the columns in $\mathbf{M}$
```{r}
apply(M, 2, mean)
```

Create a matrix $\mathbf{Mc}$, mean-centered data
```{r}
Mc <- scale(M, center = TRUE, scale = FALSE)
head(Mc, n = 10)
```

Test if $\mathbf{Mc}$ is mean-centered
```{r}
colMeans(Mc)
```

Use `sweep()` to create a mean-centered matrix of $\mathbf{M}$
```{r}
A <- sweep(M, 2, colMeans(M), "-")
head(A, 10)

head(A == Mc, 10)  # compare the result with Mc
```

Compute a vector of column maxima from $\mathbf{M}$
```{r}
maxcol_M <- apply(M, 2, max)
maxcol_M
```

Use `sweep()` to scale the columns of $\mathbf{M}$ by dividing by the column maxima
```{r}
sweep(M, 2, maxcol_M, "/")
```

Compute a matrix in which all columns of $\mathbf{M}$ scaled such that they have minimum = 0 and maximum = 1
```{r}
maxcol_M

mincol_M <- apply(M, 2, min)
mincol_M

B <- sweep(M, 2, mincol_M, "-")
B <- sweep(B, 2, maxcol_M - mincol_M, "/")
head(B, 10)
summary(B)
```

Compute the sample covariance matrix of $\mathbf{M}$
```{r}
n <- nrow(Mc)
crossprod(Mc) / (n - 1)

cov(M) # compare the result with cov(M)
```

Compute the correlation matrix of $\mathbf{M}$
```{r}
Msd <- scale(M, center = TRUE, scale = TRUE)  # create a standard matirx of M

n <- nrow(Msd)
crossprod(Msd) / (n - 1)

cor(M)  # compare the result with cor(M)
```

Write a function `dummify()` that takes a factor or a character vector, and which returns a matrix with dummy indicators.
```{r}
dummify <- function(x, all = TRUE) {
  if (!is.character(x) & !is.factor(x)) {
    stop("The input must be a character or factor vector")
  } else if (is.character(x)) {
    x <- as.factor(x)  # convert x into factor if x is a character vector
  } else if (!is.logical(all)) {
    stop("The argument 'all' must be a logical value")
  }
  
  
  
  level <- attributes(x)$levels  # create a vector of levels of x
  n <- length(level)  # number of elements in level
  
  if (all == FALSE) { 
    level <- level[-n]
    n <- n - 1
  }
  
  M <- matrix((numeric(length(x) * n)), ncol = n)
  
  for (i in 1:n) {
    M[, i] <- as.numeric(x == level[i])
  }
  
  colnames(M) <- level
  M
}
```

Test `dummify()` works
```{r}
cyl <- factor(mtcars$cyl)

CYL1 <- dummify(cyl, all = TRUE)
head(CYL1, 10)

CYL2 <- dummify(cyl, all = FALSE)
head(CYL2, 10)
```

Write a function `corsstable()` that takes two factors, and which returns a cross-table between those factors
```{r}
crosstable <- function(x, y) {
  if(!is.factor(x) | !is.factor(y)) {
    stop("The input must be two factor vectors")
  }
  
  crossprod(dummify(x), dummify(y))
}
```

Test `crostable()` works
```{r}
cyl

gear <- factor(mtcars$gear)
gear

crosstable(cyl, gear)
```